Higher-Order unification for free!

### 2024-04-16 paper/main.tex
remove unused macros
wip

\def\elpi{\proglang{elpi}}
\def\coqelpi{\proglang{coq-elpi}}
\def\lambdaprolog{\proglang{$\lambda$-prolog}}
\def\coq{\proglang{coq}}

\newcommand{\library}[1]{\textit{#1}\xspace}
\def\stdpp{\library{stdpp}}
\def\iris{\library{iris}}

\newcommand*{\acronym}[1]{\texttt{#1}\xspace}

@@

\def\ol{\acronym{ol}} % object language
\def\ml{\acronym{ml}} % meta language
\def\lf{\acronym{lf}} % logical framework
\def\ho{\acronym{ho}} % higher order
\def\Forall{$\forall$}

@@

\newcommand{\linketaM}[3]{\ensuremath{#1 \vdash #2 =_\eta #3}}
\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_{\llambda} #3}}
\newcommand{\substCell}[3]{\ensuremath{#1 \vdash #2 = #3}}
\newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}

@@

The use of meta languages facilitates the implementation of
proof systems in two key ways. The first and most well-know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.
% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
In this paper, we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

@@

\newcommand{\unifcorrectX}[4]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
   #4 \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}
\newcommand{\unifcorrect}[3]{\unifcorrectX{#1}{#2}{#3}{ \{#3_1, #3_2\} \subseteq \llambda}}

\newcommand{\unifcompleteX}[4]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #4 \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\unifcomplete}[3]{\unifcompleteX{#1}{#2}{#3}{\{#3_1, #3_2\} \subseteq \llambda}}

@@

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}\todo{give a name and reuse it for \Uo in next sections}

\noindent
Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

@@

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a detail not worth mentioning in the
previous sections).
\todo{explain fold6}

@@

\noindent
In the code above, the syntax \elpiIn{pi x y\..} is syntactic sugar for
iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
  as implemented above: \todo{reuse statement from section 2.4}
\begin{gather}
  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
  where \todo{reuse 2.4}
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

>>>>

@@

@@

\newcommand{\linketaM}[3]{\ensuremath{#1 \vdash #2 =_\eta #3}}
\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_{\llambda} #3}}
\newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}

@@

The use of a meta language facilitates the implementation of a
proof system in two key ways. The first and most well-know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.
% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
In this paper we focus on this last aspect of the implementation of
a Prolog-like proof system, precisely \emph{how to reuse the higher-order
unification procedure of the meta language in order to simulate a
higher-order logic program for the object language}.

@@

\newcommand{\unifcorrectX}[4]{\ensuremath{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
   #4 \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}}
\newcommand{\unifcorrect}[3]{\unifcorrectX{#1}{#2}{#3}{ \{#3_1, #3_2\} \subseteq \llambda}}

\newcommand{\unifcompleteX}[4]{\ensuremath{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #4 \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}}
\newcommand{\unifcomplete}[3]{\unifcompleteX{#1}{#2}{#3}{\{#3_1, #3_2\} \subseteq \llambda}}

@@

The specification of a ``good'' unification procedure restricted
to a domain $\mathcal{D}$ is the following:

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

\noindent
The meta language of choice is expected to provide 
an implementation of \Ue that is a good unification for \Ee
in \llambda.

Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

@@

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a detail not worth mentioning in the
previous sections).

@@

\noindent
In the code above, the syntax \elpiIn{pi x y\..} is syntactic sugar for
iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.
The auxiliary predicate \elpiIn{fold6} folds a predicate
with three accumulators (the memory map, the links and the substitution)
over a lists to obtain a new list and the final values of the
accumulators.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification (as per \cref{prop:good-unif})
  for \Eo in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb}
  Given the updated compilation scheme, the \Uo of~\cref{sec:founif}
  is a good unification (as per \cref{prop:good-unif})
  for \Eo in the domain $\wellb_\beta$.
\end{lemma}
